Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2015-09-03 pkg/glmertree/paper/treatment_interactions.Rnw
condensed title somewhat rephrased parts of intro (specficially mention advantages over existing methods) various cosmetic and style touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@767 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{tikz}
\linespread{1.2}
\usepackage{tabto}

@@

\usetikzlibrary{shapes,shadows,arrows,positioning}
\title{Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)}
\shorttitle{GLMM Tree}
\leftheader{GLMM Tree}
\rightheader{GLMM Tree}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

\acknowledgements{The authors would like to thank Prof. Pim Cuijpers, Prof. Jeanne Miranda, Dr. Boadie Dunlop, Prof. Rob DeRubeis, Prof. Zindel Segal, Dr. Sona Dimidjian, Prof. Steve Hollon and Erica Weitz for granting access to the dataset for the application. The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2013. The visit was supported by the Institute.}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, GLMM tree, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a generalized linear mixed-effects model for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4 and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ on the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a very generic data-analytic framework for detecting partitions in a dataset, with different model parameter estimates. The recursive partitioning in MOB can be based on a broad class of parametric models that can be fitted using M-type estimators \cite{ZeilyHoth08}, the most well-known example being the generalized linear model (GLM). Earlier, GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of datasets. In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference, due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials). In what follows, we will introduce the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning, and the generalized linear mixed-effects model (GLMM). Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree. Subsequently, we evaluate the performances of the GLMM-tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.

Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment 1 or Treatment 2. Every observation has a value on the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values on three covariates: duration of depressive symptoms prior to treatment in months (range 0-15); age in years (range 18-75); anxiety inventory total score (range 3-18).

The simulated dataset has 3 subgroups with different treatment effectiveness. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, Treatment 1 is more effective than Treatment 2: the mean of the response variable is 7 for Treatment 1, and 11 for Treatment 2. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment 1, and 9 for Treatment 2. The third subgroup consists of observations with duration $>6$. In this subgroup, Treatment 2 is more effective than Treatment 1: the mean value of the response variable is 12 for Treatment 1, and 7 for Treatment 2.

Observations were drawn from one of ten clusters, each with a different, cluster-specific (i..e, random) intercept. Data was generated such, that covariates and cluster-specific intercepts were uncorrelated. Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.

@@

To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure \ref{fig:fixedeffects}. The boxplots in Figure \ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups. There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

\begin{figure}[!ht]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm)
@
\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150). The dot for Treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

@@

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,...,J$) terminal node of the tree. As a result, in the GLM-tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':

@@

Figure \ref{fig:example_mobtree} provides an example of the GLM-tree model in Equation \ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset. By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above. However, the two subgroups in the middle result from a spurious split.

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation \ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:

@@

Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,...,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation \ref{eq:mixedeffects} with the tree from Equation \ref{eq:fixedeffects_MOB}:

@@

For estimating GLMM trees, we take the MERT approach a step further: by using a GLM tree, instead of a CART regression tree with constant fits, to estimate the fixed-effects part of the GLMM. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM-tree algorithm takes the following steps to estimate the model in Equation \ref{eq:glimmertree}:

@@

\vspace{5pt}
\noindent Step 3: Repeat steps 1 and 2 until convergence.

\vspace{5pt}
The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.

\begin{figure}[!h]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree$Tree)
@
\caption{Generalized Linear Mixed Model tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

In Figure \ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented. As can be seen, by taking into account the clustering of observations by estimating random intercepts, the spurious split involving the anxiety variable no longer appears in the tree.

@@

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure \ref{fig:modelC}. Figure \ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. In other words, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas the other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

\begin{figure}[!h]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

For each cell, 50 datasets with treatment-subgroup interactions were generated, resulting in 50 x 3 x 2 x 2 x 3 x 3 x 3 x 2 = 32,400 training datasets. For the datasets without treatment-subgroup interactions, the 6th parameter of the data-generating process had only two levels ($b$ correlated with one of the $U_k$ variables, and $b$ not correlated with any of the $U_k$ variables). Therefore, 50 x 3 x 2 x 2 x 3 x 3 x 2 x 2 = 21,600 datasets without treatment-subgroup interactions were generated.

@@

As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, -40 and 70, respectively. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between the $U_k$ variables vary according to the third facet of the simulation design described above.

@@

We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure \ref{fig:modelC}) for test observations. Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.

@@

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.

>>>>

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\linespread{1.2}
\usepackage{tabto}

@@

\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

\acknowledgements{The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application. The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014. The visit was supported by the Institute.}

@@

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and GLMMs for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM trees recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMM trees predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM trees are a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}. Specifically, this model class encompasses the generalized linear model (GLM) and GLM-based MOB trees have been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of many datasets. In such cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \cite<which is not done in other tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, we evaluate the performances of the GLMM tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.
%
Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment~1 or Treatment~2. Every observation has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values for three covariates: duration of depressive symptoms prior to treatment in months (range 0--15); age in years (range 18--75); anxiety inventory total score (range 3--18).

The simulated dataset has 3 subgroups with different treatment effectiveness. The first subgroup consists of observations with duration $\leq 6$ and anxiety $\leq 10$. In this subgroup, Treatment~1 is more effective than Treatment~2: the mean of the response variable is 7 for Treatment~1, and 11 for Treatment~2. The second subgroup consists of observations with duration $\leq 6$ and anxiety $> 10$. In this subgroup, both therapies are equally effective: the mean value of the response variable is 9 for Treatment~1, and 9 for Treatment~2. The third subgroup consists of observations with duration $>6$. In this subgroup, Treatment~2 is more effective than Treatment~1: the mean value of the response variable is 12 for Treatment~1, and 7 for Treatment~2.

Observations were drawn from one of ten clusters, each with a different, cluster-specific (i.e., random) intercept. Data was generated such that covariates and cluster-specific intercepts were uncorrelated. Also, 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, and 8\% of variance was due to cluster-specific variation.

@@

To illustrate, the GLM estimated for the artificial motivating dataset is graphically represented in Figure~\ref{fig:fixedeffects}. The boxplots in Figure~\ref{fig:fixedeffects} show the distribution of the posttreatment depression scores in both treatment groups. There seems to be little overall difference in effects of both treatments, as the slope of the regression line is nearly zero. We shall see that this does not necessarily mean that posttreatment depression score and treatment type are unrelated, as the effect of treatment may be moderated by variables not yet included in the model.

\begin{figure}[!ht]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm)
@
\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation~\ref{eq:expected_value} and~\ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

@@

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on the GLM, the result is a GLM tree, which has a local fixed-effects regression model in every $j$th ($j = 1,\dots,J$) terminal node of the tree. As a result, in the GLM tree model, the value for $\beta$ depends on terminal node $j$ in which observation $i$ `falls':

@@

Figure~\ref{fig:example_mobtree} provides an example of the GLM tree model in Equation~\ref{eq:fixedeffects_MOB}, based on the artificial motivating dataset. By using the three additional covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the treatment-subgroup interactions as described above. However, the two subgroups in the middle result from a spurious split.

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:

@@

Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}:

@@

For estimating GLMM trees, we take the MERT approach a step further: by using a GLM tree, instead of a CART regression tree with constant fits, to estimate the fixed-effects part of the GLMM. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:

@@

\vspace{5pt}
\noindent Step 3: Repeat Steps~1 and~2 until convergence.

\vspace{5pt}
The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.

\begin{figure}[!h]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree$Tree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

In Figure~\ref{fig:example_glimmertree}, the GLMM tree that was grown on the artificial motivating dataset is presented. As can be seen, by taking into account the clustering of observations by estimating random intercepts, the spurious split involving the anxiety variable no longer appears in the tree.

@@

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. In other words, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas the other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

\begin{figure}[!h]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

For each cell, 50 datasets with treatment-subgroup interactions were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets. For the datasets without treatment-subgroup interactions, the 6th parameter of the data-generating process had only two levels ($b$ correlated with one of the $U_k$ variables, and $b$ not correlated with any of the $U_k$ variables). Therefore, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 2 \times 2$ = 21,600 datasets without treatment-subgroup interactions were generated.

@@

As in \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1$, ${\mu_U}_2$, ${\mu_U}_4$, and ${\mu_U}_5$ fixed at 10, 30, $-40$, and 70, respectively. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between the $U_k$ variables vary according to the third facet of the simulation design described above.

@@

We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) for test observations. Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.

@@

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.
